---
title: "An introduction to BNPvegan"
subtitle: "Bayesian nonparametric methods for ecology"
author: "Alessandro Zito"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: BNPvegan_lib.bib
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(BNPvegan)
```

```{r, eval = FALSE}
# If the devtools R package is not already installed
# install.packages("devtools")
devtools::install_github("alessandrozito/BNPvegan")
library(BNPvegan)
```

# Introduction  

## Package description

BNPvegan is an R package that implements the fundamental tools of the Bayesian nonparametric literature on Species Sampling models. The package has been designed for ecological applications. In this spirit, we borrowed the name to mimic and extend the existing `vegan` package [@vegan]. In this tutorial we provide both a mathematical overview of the main methods available in the `BNPvegan` package, and we describe in details how to use it an how to interpret its results.

The package is constructed on two building blocks, which are highly interrelated:

* The **Species Sampling model** module -`ssm`- which runs the basic processes in the species sampling model literature [@pitman_1996], including the Dirichlet process [@Ferguson1973] and the Pitman-Yor process [@Perman_1992].  
* The **Species Discovery model** module -`sdm`- which runs the species discovery model introduced in [@Zito_2020]. 

Both functions are design to work with a standard `R` sintax, as they accept the classic methods  `summary`, `predict`, `coef` and `logLik`. Before dealing with the specifics of each block, we first describe the sample-based methods available in `BNPvegan`. 

In both cases, the main input of the package is a vector of `frequencies` representing the species counts observed in a sample. The package is endowed with two example datasets:

1. `Lepidoptera`, which the dataset of butterfly counts in [@Fisher_1943]
2. `fungalOTU`, which contains the counts for some fungal species recorded in a field experiment. 

We will use both to show the basic functioning of `ssm` and `sdm`. As a preliminary step however, we demonstrate how to obtain some simple **sample-based quantities**.

# Sample-based quantities

We hereby refer to **sample-based quantities** as those quantities which are not inherited from the the `ssm` or the `sdm` method, but can be directly derived from the vector of species `frequencies` without a specific model fit. These methods are the sample `coverage`, the `Gini` heterogeneity index and the `rarefaction` curve. 

From a mathematical perspective, let $X_1, \ldots, X_n$ be a collection of "species" tags, with $X_i$ being the $i$th species observed in the sample. We call *abundance* the total number of species observed, $n$, and *richness* the number distinct species observed in the sample, $K$. Furthermore, let $n_1, \ldots, n_K$ the frequencies with which each species appears in the sample. The three *sample-based* methods offer a way to quickly summaries the feature of the samples based on these simple quantities. 

## The sample `coverage` 
The **sample coverage** is the sum of the proportions of species that have been observed in the sample. Is is defined as 
$$
C_n = \sum_{h \in \mathcal{H}} f_h, \quad \mathcal{H} = \{\text{Inexes of the species observed among the } n \text{ data}\}
$$
where $\sum_{h=1}^H f_h = 1$ are the *species proportions* and $H$ is the total number of species in the population. Clearly, $K \leq H$, ie. the number of species discovered is less than the true number. The `BNPvegan` package offers a very simple first "estimator" for $C_n$, the old-timely Turing estimator [@Good_1953]:
$$
\hat{C}_n = 1 - \frac{m_1}{n}.
$$
with $m_1$ being the number of species observed only once. This estimator can be called in `R` with the function `coverage` as follows.

```{r}
frequencies <- as.numeric(Lepidoptera)
coverage(frequencies)
```
A first analysis then tells that roughly 99.75% of the species in the population from which `Lepidoptera` was drawn have been discovered. 

## The `Gini` heterogeneity index
The **Gini heterogeneity index** is a measure to quantify the amount of **biodiverity** present in the sample. In probabilistic terms, Gini index is the *probability that two species are different when taken at random from the population*:

$$
G = 1- \sum_{h = 1}^H f_h^2
$$
An estimator for $G$ is 
$$\hat{G} = 1 - \frac{1}{n^2}\sum_{j = 1}^K n_j^2,$$
and can be obtained by running the `Gini` command:
```{r}
Gini(frequencies)
```
In other words, the probability that two species are equal when drawn at random from the population is 0.975. This result is in line with what found with the `coverage` method

## The `rarefaction` method
While often ecologists deal with species frequencies, sometimes the species tags  $X_1, \ldots, X_n$, are observed sequentially. This allows to construct an **accumulation curve**, which plots the number of distinct species observed, $K_n$, as a function of the abundance, $n$. However, in the cases in which the data are not sequential, it is necessary to construct the so-called  **rarefaction curve**, ie. the *average accumulation curve*. This quantity is derived from a combinatorial argument and it is equal to
$$
\bar{K}_i = K - \binom{n}{i}^{-1} \sum_{j = 1}^K \binom{n - n_j}{i}.
$$
The `rarefaction` function computes precisely this quantity. Notice that the function can be computationally demanding for large $n$ and large $K$. Thus, a `verbose` option is available if one wants to monitor the process. 
```{r, cache = TRUE}
# Compute and plot the sample-based rarefaction curve
rar <- rarefaction(frequencies, verbose = TRUE)
plot(rar, type = "l", xlab = "Sample size", ylab = "Rarefaction")
```

**Note:** the `rarefaction` method performs the same task as the `rarefy` function in the `vegan` package. 


# Species Sampling models




# Species Discovery models

## Mathematical background 

- Species discovery models focus directly on the discovery of new species. Unlike Species Sampling models, they are *not* a generative model for a species distribution. 
- Species discovery models are designed to fit and predict any accumulation curve, potentially assessing its saturation level, ie. how close the curve is to convergence.

Their formulation is is the following. Let $(D_n)_{n \geq 1}$ be a sequence of *discovery indicators*, such that 

$$
D_n = \begin{cases}
1 & \text{if} \quad  X_n = ``\text{new"}\\
0 & \text{if} \quad  X_n = ``\text{observed"}
\end{cases}
$$
and trivially $D_n = 1$. Then, an accumulation curve $(K_n)_{n \geq 1}$ is defined as 
$$
K_n = \sum_{i = 1}^{n} D_i, \quad n \geq 1.
$$

The goal of a Species Discovery model is threefold:

1. **Rarefy**: obtain an in-sample estimator for the number of in-sample discoveries $\mathbb{E}(K_n)$. This is called model-based rarefaction, as it estimates a functional form for the rarefaction curve.
2. **Extrapolate**: predict the extected additional number of new species to be observed if more samples where obtained, $\mathbb{E}(K_{n+m} \mid K_n = k)$ and $m \geq 1$.
3. **Saturate**: Estimate the *sample richness* $\mathbb{E}(K_\infty\mid K_n = k)$ and the sample *saturation* $\mathbb{E}(S_n\mid K_n = k)$, with $S_n = K_n/K_\infty$.

All three goals are obtain by directly modelling the discovery probabilities linked to $D_1, \ldots, D_n$. In particular, we let $D_n \stackrel{iid}{\sim} Bernoulli(\pi_n)$, with 

$$
\pi_{n+1} = \mathbb{P}(D_{n+1} = 1) = \mathcal{S}(n;\theta), \quad \theta \in \Theta \subset \mathbb{R}^p
$$
be the discovery probability and $\mathcal{S}(n;\theta)$ a survival function such that 

* $S(0; \theta) = 1$ - the first observation is necessarily a new discovery,
* $S(n; \theta) > S(n+1; \theta)$ - the discovery probabilities descrease over time,
* $\lim_{n\to \infty}S(n;\theta) = 0$ - eventually, all species in the sample are discovered.
 
Then, any functional form that supports the three assumptions can work as a Species discovery model. Notice that under such a formulation $K_n$ is distributed as a Poisson-binomial distribution 
$$K_n \sim PB\{1,\mathcal{S}(1, \theta), \ldots,\mathcal{S}(n-1, \theta)\}.$$
This leads to two convenient estimators, namely
$$\mathbb{E}(K_n) = \sum_{i =1}^n \mathcal{S}(i-1, \theta)$$
and 
$$
\mathbb{E}(K_{n+m}\mid K_n = k) = k + \sum_{j = 0}^{n-m-1} \mathcal{S}(j, \theta).
$$

The package BNPvegan supports two type of kernel function: 

* The three-parameter log-logistic model, `LL3` (the *deault*):
$$
\mathbb{P}(D_{n+1} = 1) = \frac{\alpha\phi^n}{\alpha\phi^n + n^{1-\sigma} }
$$
with $\alpha >0$, $\sigma <1$ and $\phi \in (0,1)$. 

* The `Weibull` model:
$$
\mathbb{P}(D_{n+1} = 1) = \phi^{n^\lambda}
$$
with $\phi \in (0,1)$, $\lambda > 0$. Under both model, the value for $K_\infty = \lim_{n\to \infty} K_n$ is **guaranteed to converge**. In other words, the species discovery model will always  ensure a finite saturation. 

## The `sdm` module

The `sdm` module runs the model described above on the sample-based rarefaction curve of the specified vector of `frequencies`. The available choice for `model` are `model = "LL3"` (default) and `model = "Weibull"`. Both models are estimated via empirical Bayes by maximizing the log-likelihood associated to the collection of independent Bernoulli random variables. In what follows, we estimate the three-parameter log-logistic model on a vector of counts called `fungalOTU`. As a first step, `sdm` computes the average accumulation curve via the sample-based rarefaction formula, and then estimates the model parameters. 

```{r, cache = TRUE}
frequencies <- fungalOTU
quantile(frequencies)
```

```{r, cache = TRUE}
# Fit the species discovery model. 
# Note: verbose controls the state of the construction of the sample-based rarefaction, which is 
#       the most demanding part
fit <- sdm(frequencies, model = "LL3", verbose = TRUE)

# Summarise the output
summary(fit)
```

The quantities reported in the summary are: 

1. `Abundance`: the number of observation in the sample. 
2. `Richness`: the number of distinct species in the sample
3. `Expected species at infinity`: the asymptotic posterior species richness, ie. $\mathbb{E}(K_\infty\mid K_n = k)$
4. `Standard deviation at infinity`: the asymptotic standard deviation ie. $\{Var(K_\infty\mid K_n = k)\}^{1/2}$
5. `Expected new species to discover`: the difference between `Expected species at infinity` and `Richness`
6. `Sample saturation`: the ratio between `Richness` and `Expected species at infinity`.
7. `Parameters` the empirical Bayes estimates for the parameters of the model.

All the above quantities are easily extracted with the following commands
```{r}
# Asymptotic richness estimators
asymptotic_richness(fit)

# Parameters
coef(fit)

# Loglikelihood
logLik(fit)
```

Once the model is estimated, it is very easy to plot both the in-sample rarefaction and to obtain the out of sample prediction. First, let's start with a plot. 


```{r}
# Plot the model-based rarefaction curve vs the average accumulation curve
plot(fit, type = "rarefaction")
```


# References

