---
title: "Short tutorial"
author: "Alessandro Zito"
date: "24/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BNPvegan)
library(vegan)
library(tidyverse)
```

### PRELIMINARY STEP - Code up the multinomial model rarefaction and extraoplations
```{r}
rarefy_multinomial <- function(abundances, m){
  S_obs <- length(abundances)
  freq <- freq_abundance(abundances, r = 1:sum(abundances))
  n <- sum(abundances)
  alpha_km <- c(exp(lchoose(n - 1:(n-m), m) - lchoose(n, m)), rep(0, m))
  return(S_obs - sum(alpha_km * freq))
}
rarefy_multinomial <- Vectorize(rarefy_multinomial, vectorize.args = "m")

rarefy_multinomial(abundances, m = 1:10)
c(rarefy(abundances, sample = 1:10))
rarefaction(abundances, verbose = FALSE)[1:10]
```

Shen et al (2004) extrapolation estimator:
$$
\tilde{S}_{ind}(n + m) = K + \hat{f}_0\Bigg[1- \exp \bigg(-\frac{m}{n}\frac{f_1}{\hat{f}_0}\bigg)\Bigg]
$$
where $\hat{f}_0$ is the asymptotic number of undiscovered species estimator and $f_1$ is the number of singletons in the abundances vector.
 
```{r}
# Extrapolation estimator based in Shen et al 2004
extrapolate_ShenChao <- function(abundances, f0, m){
  f1 <- sum(abundances == 1)
  S_obs <- length(abundances)
  return(S_obs + f0*(1- exp(- m/n * f1/f0)))
}

extrapolate_ShenChao <- Vectorize(extrapolate_ShenChao, vectorize.args = "m")
rich <- estimateR(abundances)
f0 <- unname(rich[4] - rich[1])
extrapolate_ShenChao(abundances, f0, m = 500:505)
```

Chao et al (2009) additional samples estimators
$$
m_g = \frac{nf_1}{2f_2}\log\Bigg[\frac{\hat{f}_0}{(1-g)S_{est}} \Bigg], \quad S_{obs}/S_{est} < g<1
$$
what if there are no doubletons?
```{r}
targetSat_Chao <- function(abundances, g, f0){
  f1 <- sum(abundances == 1)
  f2 <- sum(abundances == 2)
  n <- sum(abundances)
  S_obs <- length(abundances)
  S_est <- f0 + S_obs
  if(S_obs/S_est < g ){
    return(n*f1/(2*f2) *log(f0/((1-g)*S_est)))
  } else{
    cat("g too low. Saturation is", S_obs/S_est, "\n")
  }
}
```



## SAMPLE-BASED quantities
```{r}
abundances <- as.numeric(Lepidoptera)
```


```{r}
abundances <- fungalOTU
table(abundances)
```

```{r}
##### Sample-based quantities. No modelling assumptions. 

# Coverage method  --> fraction of species observed in the whole population
coverage(abundances)

# Gini heterogeneity  ---> prob. that two randomly drawn species from the pop. are equal.
Gini(abundances)

# Rarefaction ---> average accumulation curve
rar <- rarefaction(abundances, verbose = TRUE) 
plot(rar, type = "l")

# In the vegan package, this is called by running rarecurve
rar_vegan <- rarecurve(data.frame(t(as.matrix(abundances))), step = 1)[[1]]

head(rar, 10)
head(unname(rar_vegan), 10)
```

## Species sampling models
```{r}
# Random quantities generation
counts <- rPY(size = 1000, alpha = 1, sigma = 0.1)
print(counts)

rDP(size = 1000, alpha = 1)
```


```{r}
# Species sampling models

# Dirichlet 
fitDP <- ssm(abundances, model = "DP")
summary(fitDP)

# Pitman-Yor
fitPY <- ssm(abundances, model = "PY")
summary(fitPY)

# Compare the loglikelihood 
logLik(fitDP)
logLik(fitPY)
```

```{r}
coef(fitPY)
logLik(fitPY)
```

```{r}
# Summarize the quantities in the Pitman-Yor

# Under r = 1 means that you are finding the expected number of singletons in the model
freq_abundance(abundances, r = 1:10)
freq_abundance(fitPY, r = 1:10)
plot(fitPY, type = "freq")
```

```{r}
## Model-based coverage
coverage(fitPY)
plot(fitPY, type = "coverage")
rcoverage(fitPY, n_samples = 10)
```

```{r}
# Gini diversity for the Pitman-Yor
Gini(fitPY)
Gini(abundances)  # Can be sampled from. Need a stick breaking procedure. 

hist(rGini(fitPY, n_samples = 500))
```

```{r}
# Model-based rarefaction 
plot(fitPY, type = "rarefaction", plot_sample = TRUE, verbose = FALSE)
plot(fitDP, type = "rarefaction", plot_sample = TRUE, verbose = FALSE)
rarPY <- rarefaction(fitPY)
head(rarPY)
```
```{r}
# Model-based extrapolation
extrapolation(fitPY, m = 10000)
plot(fitPY, type = "extrapolation", m = 1000, verbose = FALSE, plot_sample = TRUE)
```

## Species discovery models
```{r}
## Run a species discovery model on the same dataset
abundances = Lepidoptera
fit_sdm <- sdm(abundances, model = "LL3", verbose = FALSE)
summary(fit_sdm)
logLik(fit_sdm)
coef(fit_sdm)
```

```{r}
# Usual model-based quantities
coverage(fit_sdm)

# rarefaction
rar_sdm <- rarefaction(fit_sdm)
head(rar_sdm)
plot(fit_sdm, type = "rarefaction")

# Extrapolation
extrapolation(fit_sdm, m = 1000)
plot(fit_sdm, type = "extrapolation", m = 10000)

# Asymptotic richness
asym_richness(fit_sdm)
```

```{r}
# The posterior species richness is actually a random variable!
postR <- sample_Kinf(fit_sdm, n_samples = 1000)
hist(postR)
```
```{r}
# Saturation
saturation(fit_sdm, method = "approximate")

# Random samples from the saturation
sat <- saturation(fit_sdm, method = "montecarlo", n_samples = 1000)
hist(sat, breaks = 20)
```

```{r}
# Target method to acheive a given level of saturation
saturation(fit_sdm)
saturation(fit_sdm, method = "target", target = 0.95)
```
 
Compared to Chao et al 2009: 
```{r}
rich <- estimateR(abundances)
# Under ACE
f0 <- unname(rich[4]-rich[1])
targetSat_Chao(abundances, g = 0.95, f0)

# Under Chao1
f0 <- unname(rich[2]-rich[1])
targetSat_Chao(abundances, g = 0.95, f0)
```


```{r}
# Comparison of extrapolation methods
abundances <- as.numeric(fungalOTU)
m <- sum(abundances)

# Individual based rarefaction
rar <- rarefaction(abundances, verbose = FALSE)

# Dirichlet process
fit_DP <- ssm(abundances, model = "DP")
rar_DP <- rarefaction(fit_DP)
extr_DP <- extrapolation(fit_DP, 1:m)

# Pitman-Yor process
fit_PY <- ssm(abundances, model = "PY")
rar_PY <- rarefaction(fit_PY)
extr_PY <- extrapolation(fit_PY, 1:m)

# LL3 sdm
fit_sdm <- sdm(abundances, verbose = FALSE)
rar_sdm <- rarefaction(fit_sdm)
extr_sdm <- extrapolation(fit_sdm, 1:m)

# Chao extrapolation
rich <- estimateR(abundances)
f0 <- unname(rich[4]-rich[1])  # use the ACE estimator
extr_chao <- extrapolate_ShenChao(abundances, f0, m = 1:m)
```


```{r}
# Everything in a dataframe
cutoff <- sum(abundances)
df <- data.frame( "n" = c(1:(cutoff + m)),
                 "rar"= c(rar, rep(NA, m)),
                 "DP" = c(rar_DP, extr_DP),
                 "PY" = c(rar_PY, extr_PY), 
                 "sdm" = c(rar_sdm, extr_sdm), 
                 "Chao" = c(rep(NA, cutoff), extr_chao))

n_points = 200
if (nrow(df) > n_points) {
  seqX <- 1:nrow(df)
  seqY <- split(seqX, sort(seqX %% n_points))
  df <- df[unlist(lapply(seqY, function(a) utils::tail(a, 1))), ]
}

ggplot(data = df[df$n>1000,])+
  
  geom_line(aes(x = n, y = DP, color = "DP"), na.rm = TRUE, size = 1) +
  geom_line(aes(x = n, y = PY, color = "PY"), na.rm = TRUE, size = 1, linetype = "dashed") +
  geom_line(aes(x = n, y = sdm, color = "SDM"), na.rm = TRUE, size = 1) +
  geom_line(aes(x = n, y = Chao, color = "Chao"), na.rm = TRUE, size = 1, linetype = "dotdash") +
  geom_segment(x = cutoff, xend = cutoff, y = 0, yend = Inf, linetype = "dashed")+
  geom_point(aes(x = n, y = rar), na.rm = TRUE, size = 1, shape = 1) +
  theme_bw()
```
```{r}
fit_PY <- ssm(fungalOTU, model = "PY")
```



