---
title: "An introduction to BNPvegan"
subtitle: "Bayesian nonparametric methods for ecology"
author: 
  - Alessandro Zito^[Duke University, alessandro.zito@duke.edu]
  - Tommaso Rigon^[University of Milano-Bicocca, tommaso.rigon@unimib.it]
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: BNPvegan_lib.bib
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(BNPvegan)
```


# Introduction  

## Installation

```{r, eval = FALSE}
# If the devtools R package is not already installed
# install.packages("devtools")
devtools::install_github("alessandrozito/BNPvegan")
library(BNPvegan)
```


## Package description

[`BNPvegan`](https://github.com/alessandrozito/BNPvegan) is an R package that implements some fundamental tools of the Bayesian nonparametric literature on Species Sampling models. The code is available at this [github repository](https://github.com/alessandrozito/BNPvegan). The package has been designed for ecological applications. In this spirit, we chose to name it to mimic and extend the existing `vegan` package [@vegan]. In this tutorial we provide both a mathematical overview of the main methods available in `BNPvegan`, and we describe in details how to use it and how to interpret its results.

The package is constructed on two building blocks, which are highly interrelated:

* The **Species Sampling model** module -`ssm`- which runs the basic processes in the species sampling model literature [@pitman_1996], namely the Dirichlet process [@Ferguson1973] and the Pitman-Yor process [@Perman_1992].  
* The **Species Discovery model** module -`sdm`- which runs the species discovery model introduced in @Zito_2020. 

Both functions are designed to work with a standard `R` syntax, as their output can be accessed via the classic methods `summary`, `plot`, `coef` and `logLik`. 

In both cases, the main input of the package is a vector of `abundances` representing the species counts observed in a sample. The package is endowed with two example datasets:

1. `Lepidoptera`, which is the dataset of butterfly counts in @Fisher_1943
2. `fungalOTU`, which contains the counts for some fungal species recorded in a field experiment. 

We will use both to show the basic functioning of `ssm` and `sdm`. As a preliminary step, however, we show how to obtain some simple **sample-based quantities**.

# Sample-based quantities

We hereby refer to **sample-based quantities** as those indicators which are not inherited from the the `ssm` or the `sdm` method, but can be directly derived from the vector of species `abundances` without a specific model fit. These methods are the sample `coverage`, the `Gini` heterogeneity index and the `rarefaction` curve. 

From a mathematical perspective, let $X_1, \ldots, X_n$ be a collection of "species" tags, with $X_i$ being the $i$th species observed in the sample. We call *abundance* the total number of species observed, $n$, and *richness* the number distinct species observed in the sample, $K$. Furthermore, let $n_1, \ldots, n_K$ the abundances with which each species appears in the sample. The three *sample-based* methods offer a way to quickly summarize vector of `abundances`. We describe them in details one by one.

## The sample `coverage` 

The **sample coverage** is the sum of the proportions of species that have been observed in the sample. Is is defined as 
$$
C_n = \sum_{h \in \mathcal{H}} f_h, \quad \mathcal{H} = \{\text{"Indexes of the species observed among the } n \text{ data"}\}
$$
where $\sum_{h=1}^H f_h = 1$ are the *species proportions* and $H$ is the total number of species in the population. Clearly, $K \leq H$, ie. the number of species discovered is less than the true number. The `BNPvegan` package offers a very simple first "estimator" for $C_n$, the old-timely Turing estimator [@Good_1953]:
$$
\hat{C}_n = 1 - \frac{m_1}{n}.
$$
with $m_1$ being the number of species observed only once. This estimator can be called in `R` with the function `coverage` as follows.

```{r}
# Turing estimator of the sample coverage
abundances <- as.numeric(Lepidoptera)
coverage(abundances)
```
This tells that roughly 99.75% of the species in the population from which `Lepidoptera` was drawn have been discovered. 

## The `Gini` heterogeneity index
The **Gini heterogeneity index** is a measure to quantify the amount of **biodiversity** present in the sample. In probabilistic terms, the Gini index is the *probability that two species are different when taken at random from the population*:

$$
G = 1- \sum_{h = 1}^H f_h^2.
$$
An estimator for $G$ is 
$$\hat{G} = 1 - \frac{1}{n^2}\sum_{j = 1}^K n_j^2,$$
and can be obtained by running the `Gini` command:
```{r}
# Sample-based Gini index
Gini(abundances)
```
In other words, the probability that two species are equal when drawn at random from the population is 0.975. This result is in line with what found with the `coverage` method. 

## The `rarefaction` method

While often ecologists deal with species frequencies, sometimes the species tags  $X_1, \ldots, X_n$, are observed sequentially. This allows to construct an **accumulation curve**, which plots the number of distinct species observed, $K_n$, as a function of the abundance, $n$. However, in the cases in which the data are not sequential, one can cope for the ordering issue by constructing the so-called  **rarefaction curve**, ie. the *average accumulation curve*. This quantity is derived from a combinatorial argument and it is equal to
$$
\bar{K}_i = K - \binom{n}{i}^{-1} \sum_{j = 1}^K \binom{n - n_j}{i},
$$
for $i = 1, \ldots, n$. To compute it, it is sufficient to call the function `rarefaction` on a vector of `frequencies`. Notice that the function can be computationally demanding for large $n$ and large $K$. Thus, a `verbose` option is available if one wants to monitor the process. 

```{r, cache = TRUE}
# Compute and plot the sample-based rarefaction curve
rar <- rarefaction(abundances, verbose = TRUE)
plot(rar, type = "l", xlab = "Sample size", ylab = "Rarefaction")
```

**Note:** the `rarefaction` method performs the same task as the `rarecurve` function in the `vegan` package.  

While being relatively quick and easy to interpret, the sample-based quantities are not the product of a model. As such, there is no way to include uncertainty quantification in their estimates. In the next sections, however, we show how the methods `ssm` and `sdm` can solve for this issue.

# Species Sampling models

## Sequential allocation scheme

Species sampling models [@pitman_1996] are a class of **generative models** for species abundances which belong to the Bayesian nonprametric literature. The most common species sampling models are the Dirichlet process, `"DP"` [@Ferguson1973] and the Pitman-Yor process, `"PY"` [@Perman_1992]. For the purpose of this guide, we will only provide an overview of the basic properties of both models. For an extensive overview on how species sampling model work, see @DeBlasi_2015.

Both the Dirichlet and the Pitman-Yor process generate species abundances through a sequential prediction scheme. Let $X_{1}, \ldots, X_{n}$ be a sequence of individuals observed, $K_n$ the total number of distinct species observed and  $X_{1}^*, \ldots, X_{K_n}^*$  the distinct species labels. In other words, $X_{j}^*$ is the $j$th species label observed in among the first $n$ observations. Then, the species of $X_{n+1}$ is determined by the following prediction rules:

* In the **Dirichlet process**: $$X_{n+1} \mid X_1, \ldots, X_n \sim \frac{\alpha}{\alpha + n}(\text{"new species"}) + \sum_{j=1}^{K_n} \frac{n_j}{\alpha + n}\delta_{X_j^*}$$
with $\alpha >0$;
* In the **Pitman-Yor process**: 
$$
X_{n+1} \mid X_1, \ldots, X_n \sim \frac{\alpha + \sigma K_{n}}{\alpha + n}(\text{"new species"}) + \sum_{j=1}^{K_n} \frac{n_j - \sigma}{\alpha + n}\delta_{X_j^*}
$$ 
with $\alpha >-\sigma$ and $\sigma \in [0, 1)$.

The parameters $\alpha$ and $\sigma$ govern the allocation scheme. In particular, it is easy to see that the Pitman-Yor process degenerates to the Dirichlet process when $\sigma = 0$. The interpretation of both equations is rather simple: the species observed at time $n+1$ can either be of a new type or of a previously observed one. In particular, the probability of observing an "old" species is increasing in the abundance of that particular species. The package `BNPvegan` contains a random sampler for both processes, which returns a vector of species abundances. 

```{r}
set.seed(1)
n <- 1000
# Random sample from a Dirichlet process
rDP(n, alpha = 1)  
# Random sample from a Pitman-Yor process 
rPY(n, alpha = 1, sigma = 0.3)  
```

In the following subsection, we show how to perform inference on the parameters $\alpha$ and $\sigma$ via the `ssm` module. 

## The `ssm` module
In what follows, we describe the functioning of the `ssm` module by estimating a Pitman-Yor on the `Lepidoptera` dataset. All the functions are equivalent for the Dirichlet process. 

The likelihood function of the Pitman-Yor process is given by the following quantity (known as exchangeable partition probability function):
$$
\mathcal{L}(\alpha, \sigma \mid X_1, \ldots, X_n) = \frac{\prod_{j=1}^{K_n-1}(\alpha + j\sigma)}{(\alpha + 1)_{n-1}}\prod_{j=1}^{K_n}(1-\sigma)_{n_j - 1}.
$$
To estimate $\alpha$ and $\sigma$, the `ssm` function adopts an empirical Bayes procedure and performs a simple off-theshelf numerical optimization of $\mathcal{L}(\alpha, \sigma \mid X_1, \ldots, X_n)$ via `nlminb`. The commands are the following.

```{r}
# Estimate a Pitman-Yor model
abundances <- as.numeric(Lepidoptera)
fit <- ssm(abundances, model = "PY")  # set model = "DP" for the Dirichlet process
summary(fit)  # Summarize the output
```

The `summary` function applied to an object of class `ssm` returns the following quantities:

1. `Abundance`: the number of observation in the sample. 
2. `Richness`: the number of distinct species in the sample
3. `Estimated sample coverage`: model-based sample coverage
4. `Posterior Gini diversity`: model-based Gini diversity
4. `Expected species after additional xx samples`: model-based extrapolation
5. `New expected species after additional xxx samples:`: the difference between `Expected species after additional xx samples` and `Richness`
6. `Expected species at infinity`: aysymptotic species richness under the model
7. `Parameters` the empirical Bayes estimates for the parameters of the model.

To access the parameters and the loglikelihood, do
```{r}
coef(fit)
logLik(fit)
```

Notice that in this particular case, $\sigma$ is estimated to be 0 (ie. the Pitman-Yor degenerates to the Dirichlet process). To assess the goodness-of-fit of the `ssm` function, we plot the observed frequency-of-abundances against the model-based ones.

```{r}
# Frequency-of-frequencies plot
plot(fit, type = "freq")
```

Such a plot can be obtained by looking at the model-based frequency-of-abundances, which are available in closed form for the Pitman-Yor process [@Favaro_2013]. In particular, for a given $r = 1, \ldots, n$, the number of species appearing $r$ times is is equal to

$$
\mathbb{E}(M_r) = \frac{\alpha}{(\alpha)_n}\binom{n}{r}(1-\sigma)_{r-1}(\alpha + \sigma)_{n-r}.
$$
To obtain it, just use the following method

```{r}
# Frequency-of-abundances estimate under Pitman-Yor model
freq_abundance(abundances, r = 1:10)
freq_abundance(fit, r = 1:10)
```

The `ssm` module offers an alternative to the sample-based estimators described in the previous section. In particular, it is endowed with a **model-based** `coverage`, `Gini` diversity and `rarefaction` function. Moreover, due to the sequential urn scheme characteristic, it also allows to **extrapolate** the rarefaction curve via the command `extrapolation`, ie. predict the additional number of distinct species if additional samples where observed. Let's see the one by one.

### Model-based `coverage`

In a Pitman-Yor model, the distribution of the posterior sample coverage is 
$$C_n \mid X_1, \ldots, X_n \sim Beta(n-\sigma K_n, \alpha + \sigma K_n)$$
and the posterior mean is 
$$\mathbb{E}(C_n \mid X_1, \ldots, X_n) = \mathbb{P}(X_{n+1} = \text{old species} \mid X_1, \ldots, X_n) = \frac{n-\sigma K_n}{\alpha + n}$$
The advantage of this result is that it allows to obtain **uncertainty quantification on the coverage**.  To see this, just run the following commands:

```{r, cache = TRUE}
# Estimate for the mean coverage under a Pitman-Yor model
coverage(fit)
# Random samples from the posterior distribution of the coverage
coverage_samples <- rcoverage(fit) 
head(coverage_samples)
# Plot the posterior distribution of the coverage
plot(fit, type = "coverage")
```

### Posterior `Gini` index

The Pitman-Yor process admits a closed for estimator for the Gini index. In particular, the **a priori** Bayesian estimate for the Gini index is 
$$\mathbb{E}(G) = \frac{\alpha + \sigma}{\alpha + 1},$$ while the **a posteriori** estimate is
$$
\mathbb{E}(G\mid X_1\ldots, X_n)  = 1- \frac{1}{(\alpha + n)_2}\Big\{(1 -\sigma)(\alpha + K_n\sigma) + \sum_{j=1}^K(n_j - \sigma)_2\Big\} 
$$
where $(a)_n = \Gamma(a + n)/\Gamma(a)$ is the Pochhammer symbol. To obtain the estimate for the expected posterior Gini index, run the `Gini` function.

```{r}
# Expected nodel-based Gini inidex
Gini(fit)
```
Like in the `coverage` method, we can also obtain posterior samples for the Gini index. This might be computationally slow, as obtaining such samples requires running a stick breaking construction. 
```{r}
# Random samplers for the Gini coefficient
set.seed(1)
Gini_samples <- rGini(fit, n_samples = 500)
hist(Gini_samples, main = "Posterior Gini index")
```


### Model-based `rarefaction` and `extrapolation`. 

Both the Dirichlet and the Pitman-Yor process assume that the observations $X_1, \ldots, X_n$ are **exchangeable**. In other words, the order in which they appear does not matter. Moreover, we can show that the rarefaction curve in the Pitman-Yor process is a Markov process such that $K_1 = 1$ and 

$$
K_{n+1} \mid K_n = K_{n-1} + D_n, \quad (D_n \mid K_n)\sim \text{Bern}\Big(\frac{\alpha + \sigma K_n}{\alpha + n}\Big)
$$

Moreover, the associated Bayesian estimate for the rarefaction curve is
$$\mathbb{E}(K_n) = \frac{\alpha}{\sigma}\Big\{ \frac{(\alpha + \sigma)_n}{(\alpha)_n} - 1\Big\}$$
The numerical values and the associated plot for the rarefaction curve are obtained as follows.

```{r}
# Model-based rarefaction curve
rar <- rarefaction(fit)
head(rar)
tail(rar)
```
 
```{r, cache = TRUE}
# Plot the rarefaction curve
plot(fit, type = "rarefaction", verbose = FALSE, plot_sample = TRUE)
```


A convenient property of the Pitman-Yor process is the fact that the prediction of additional number of distinct species if $m$ new samples where observed is available in closed form. In particular, following @Favaro_2009, we have that 
$$
\mathbb{E}(K_m^{(n)} \mid K_n = k) = \Big(k + \frac{\alpha}{\sigma}\Big)\Big\{\frac{(\alpha + n + \sigma)_m}{(\alpha + n)_m} - 1\Big\}
$$
where $K_m^{(n)}$ is the number of new species observed under additional $m$ samples after $n$ individuals have been observed. The `extrapolation` function then computes the quantity $k + \mathbb{E}(K_m^{(n)} \mid K_n = k)$ for any given value of the parameter $m$. In particular:

```{r}
# Obtain the extrapolation estimates for m additional samples
extr <- extrapolation(fit, m = c(1:1000))
head(extr)
tail(extr)
```

```{r, cache = TRUE}
# Plot the extrapolation curve
plot(fit, type = "extrapolation", m = 3000, plot_sample = TRUE, verbose = FALSE)
```


**A final note**: in the Pitman-Yor process, the growth rate of the number of clusters $K_n$ is $\mathcal{O}(n^\sigma)$, while in the Dirichlet process it is $\mathcal{O}(\alpha \log n)$. In other words, **both models assume an infinite number of species asymptotically**. Thus, they cannot be used to assess the sample saturation. Dealing with converging accumulation curves is the purpose of the `sdm` module. 

# Species Discovery models

## Background 

- Species discovery models focus directly on the discovery of new species. Unlike Species Sampling models, they are *not* a generative model for a species abundances. 
- Species discovery models are designed to fit and predict any accumulation curve, potentially assessing its saturation level, ie. how close the curve is to convergence.
- Their formulation involves a **direct** modelling of the accumulation curve obtained from the sample. 

Let $(D_n)_{n \geq 1}$ be a sequence of *discovery indicators*, such that 

$$
D_n = \begin{cases}
1 & \text{if} \quad  X_n = ``\text{new"}\\
0 & \text{if} \quad  X_n = ``\text{observed"}
\end{cases}
$$
and trivially $D_1 = 1$. Then, an accumulation curve $(K_n)_{n \geq 1}$ is defined as 
$$
K_n = \sum_{i = 1}^{n} D_i, \quad n \geq 1.
$$

The goal of a Species Discovery model is threefold:

1. **Rarefy**: obtain an in-sample estimator for the number of in-sample discoveries $\mathbb{E}(K_n)$. This is a model-based rarefaction, as it estimates a functional form for the rarefaction curve.
2. **Extrapolate**: predict the expected additional number of new species to be observed if more samples where obtained, $\mathbb{E}(K_{n+m} \mid K_n = k)$ and $m \geq 1$.
3. **Saturate**: Estimate the *sample richness* $\mathbb{E}(K_\infty\mid K_n = k)$ and the sample *saturation* $\mathbb{E}(S_n\mid K_n = k)$, with $S_n = K_n/K_\infty$.

All three goals are obtain by directly modelling the discovery probabilities linked to $D_1, \ldots, D_n$. In particular, we let $D_n \stackrel{iid}{\sim} Bernoulli(\pi_n)$, with 

$$
\pi_{n+1} = \mathbb{P}(D_{n+1} = 1) = \mathcal{S}(n;\theta), \quad \theta \in \Theta \subset \mathbb{R}^p
$$
be the discovery probability and $\mathcal{S}(n;\theta)$ a survival function such that 

* $S(0; \theta) = 1$ - the first observation is necessarily a new discovery,
* $S(n; \theta) > S(n+1; \theta)$ - the discovery probabilities decrease over time,
* $\lim_{n\to \infty}S(n;\theta) = 0$ - eventually, all species in the sample are discovered.

Then, any functional form that supports the three assumptions can work as a species discovery model. The package `BNPvegan` supports two types of survival functions: 

* The three-parameter log-logistic model, `LL3` (the *default*):
$$
\mathbb{P}(D_{n+1} = 1) = \frac{\alpha\phi^n}{\alpha\phi^n + n^{1-\sigma} }
$$
with $\alpha >0$, $\sigma <1$ and $\phi \in (0,1)$. Notice that for $\phi = 1$ and $\sigma = 0$, `LL3` has the same discovery probability of the Dirichlet process .

* The `Weibull` model:
$$
\mathbb{P}(D_{n+1} = 1) = \phi^{n^\lambda}
$$
with $\phi \in (0,1)$, $\lambda > 0$. 

Under both model, the value for $K_\infty = \lim_{n\to \infty} K_n$ is **guaranteed to converge**. In other words, the species discovery model will always ensure a finite saturation. 

## The `sdm` module

The `sdm` module runs the model described above on the sample-based rarefaction curve of the specified vector of `frequencies`. The available choice for `model` are `model = "LL3"` (default) and `model = "Weibull"`. Both models are estimated via empirical Bayes by maximizing the log-likelihood associated to the collection of independent Bernoulli random variables. This quantity is equal to 
$$
\mathcal{L}(\theta\mid D_1, \ldots, D_n) = \prod_{i = 2}^n  \mathcal{S}(i-1;\theta)^{D_n}(1-\mathcal{S}(i-1;\theta))^{1-D_n}
$$
where the degenerate case $i= 1$ is trivially excluded. As already mentioned above, it is important to notice that often the sequence of species discovery is not available. In the sample based-methods, this problem is solved by computing the **rarefaction** curve from the vector of given `frequencies`, while in the `ssm` module the **exchangeability** assumption implies that the ordering is irrelevant. In the `sdm` case however, the order of the discoveries does matter. To solve this issue, the module maximizes the following likelihood:
$$
\mathcal{L}(\theta\mid \tilde{D}_1, \ldots, \tilde{D}_n) = \prod_{i = 2}^n  \mathcal{S}(i-1;\theta)^{\tilde{D}_n}(1-\mathcal{S}(i-1;\theta))^{1-\tilde{D}_n}
$$
where $\tilde{D}_i = \bar{K}_{i+1}- \bar{K}_i$ and $\bar{K}_i$ is derived from the average sample-based rarefaction curve. Thus, `sdm` works in three steps 

1. Compute the  sample-based rarefaction curve $\hat{K}_i$ for $i = 1, \ldots, n$
2. Compute the differences and $\tilde{D}_i = \hat{K}_{i+1}- \hat{K}_i$ with $\tilde{D}_i = 1$
3. Obtain the empirical Bayes estimates for $\theta$ via constrained maximization of the likelihood $\mathcal{L}(\theta\mid \tilde{D}_1, \ldots, \tilde{D}_n)$.

In the case of `model = "LL3`, the maximization of the likelihood can be obtained with a simple constrained logistic regression (package `glmnet`). This is makes the `sdm` function particularly efficient

```{r, cache = TRUE}
# Fit the species discovery model. 
# abundances <- fungalOTU
abundances <- Lepidoptera

# Note: verbose controls the state of the construction of the sample-based rarefaction, which is 
#       the most demanding part
fit_sdm <- sdm(abundances, model = "LL3", verbose = FALSE)

# Summarise the output
summary(fit_sdm)
```

The quantities reported in the summary are: 

1. `Abundance`: the number of observation in the sample. 
2. `Richness`: the number of distinct species in the sample
3. `Estimated sample coverage`: model-based sample coverage
4. `Expected species after xx additional samples`: model-based rarefaction
5. `Expected new species after xx additional samples`: difference between `Expected species after xx additional samples` and `Richness`
6. `Expected species at infinity`: the asymptotic posterior species richness, ie. $\mathbb{E}(K_\infty\mid K_n = k)$
7. `Standard deviation at infinity`: the asymptotic standard deviation ie. $\{Var(K_\infty\mid K_n = k)\}^{1/2}$
8. `Expected new species to discover`: the difference between `Expected species at infinity` and `Richness`
9. `Sample saturation`: the ratio between `Richness` and `Expected species at infinity`.
10. `Parameters` the empirical Bayes estimates for the parameters of the model.

The parameters are easily extracted with the following commands

```{r}
# Parameters
coef(fit_sdm)

# Loglikelihood
logLik(fit_sdm)
```

## A `rarefaction` and an `extrapolation` estimator
The main advantage of the `sdm` framework is that it is naturally endowed with a  `rarefaction` and an `extrapolation` estimator. Indeed, we can prove that the number of distinct species $K_n$ is distributed as a Poisson-binomial distribution,
$$K_n \sim PB\{1,\mathcal{S}(1, \theta), \ldots,\mathcal{S}(n-1, \theta)\},$$
with mean
$$\mathbb{E}(K_n) = \sum_{i =1}^n \mathcal{S}(i-1, \theta).$$
The above quantity can be obtained by using the `rarefaction` method.

```{r, cache = TRUE}
# Extract the model-based rarefaction curve for sdm
rar_sdm <- rarefaction(fit_sdm)
head(rar_sdm)
tail(rar_sdm)
# Plot the model-based and the sample-based rarefaction
plot(fit_sdm, type ="rarefaction") 
```

As for extrapolation, it can be easily proved that for any $m\geq 1$, 

$$
\mathbb{E}(K_{n+m}\mid K_n = k) = k + \sum_{j = 0}^{n-m-1} \mathcal{S}(j, \theta).
$$

Again, the method works in the same way as in the `ssm` module
```{r, cache = TRUE}
# Extrapolate in sdm
extr_sdm <- extrapolation(fit_sdm, m = 1:1000)
head(extr_sdm)
tail(extr_sdm)

# Plot in sdm
plot(fit_sdm, type = "extrapolation", m = 10000)
```

## Saturation and asymptotic richness methods
The final feature of `sdm` is the fact that it provides an estimate for the asymptotic species richness, which is the total number of distinct species present in the population. The estimator is equal to 
$$\mathbb{E}(K_\infty \mid K_n = k) = k + \sum_{j = n}^\infty S(j; \theta) <\infty.$$
Moreover, $K_\inf\mid K_n = k$ is a random variable. To simulate from it, run
```{r, cache = TRUE}
### Simulate from the posterior distribution of the species richness
set.seed(1)
Kinf <- sample_Kinf(fit_sdm, size = 1000)
plot(density(Kinf), main = "Posterior species richness")
```
By sampling from the posterior species richness we can also assess the **saturation** level, which is the fraction of species discovered over the asymptotic estimate, namely
$$S_n\mid K_n = k = \frac{k}{K_\infty}.$$
As computing the expected value for this quantity can be demanding, `sdm` gives the possibility to compute $k/\mathbb{E}(K_\infty \mid K_n = k)$, which is a first approximation. This can be achieved my specifying `method = "approximate"` (the default). Otherwise, if one wants to carry uncertainty quantification on the saturation level, set `method = "montecarlo"` and specify the number of samples `n_samples` to draw. This last option can be slow. See the following code:

```{r, cache = TRUE}
# Asymptotic richness estimators
asym_richness(fit_sdm)
# Approximate saturation
saturation(fit_sdm, method = "approximate")
# Samples from the posterior saturation
set.seed(1)
post_sat <- saturation(fit_sdm, method = "montecarlo", n_samples = 500)
plot(density(post_sat), main = "Posterior saturation")
```

Finally, we can compute the approximate number of additional samples to draw to reach a desired level of saturation. Can be obtained by specifying `method = "target"` and a desired `target`. Note that `target` has to be between 0 and 0.999 for stability reasons.
```{r}
# Additional samples to reach a desired level of saturation
saturation(fit_sdm, method = "target", target = 0.9)
saturation(fit_sdm, method = "target", target = 0.99)
```

# References



